Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-06 paper/ms.tex
update ms

\paragraph{Cep-Her}
The $\approx$40\,Myr old Cep-Her complex currently has its most
constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs, and it contains Kepler-1627Ab,
Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
\citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
may be older by $\approx$50\% (\citealt{Bouma_2022b};
Kerr~et.al.~in preparation).  The planets themselves are all on
close-in orbits (5-25\,days), with sizes from 2-4\,$R_\oplus$.  The
rotation, lithium, and cluster ages all agree within 1$\sigma$ for
Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the nominal cluster age for the RSG-5 subgroup.

\paragraph{MELANGE-3}
\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
the Kepler field and derived its age using a combination of rotation,
lithium, and the isochronal pre-main-sequence.  Kepler-970 and
Kepler-1928 harbor two transiting mini-Neptunes inside this
association \citep{2022AJ....164...88B}.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and -1928 respectively.  The former rotation-based age
being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.

\paragraph{Theia~520}
Kepler-968 and Kepler-52 both have rotation-based ages of
330-350\,Myr.  These stars were reported by
\citet{2019AJ....158..122K} to be members of an association called Theia~520
based on their spatial and kinematic correlations with nearby stars.
The core of this structure has been independently called UBC-1
\citep{2018A&A...618A..59C}; since both Kepler planet hosts
are in diffuse regions not covered in this latter study, we use the name from
\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting, Theia~520
is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
An independent analysis of the cluster's core by
\citet{2024A&A...681A..13F} using g-mode asteroseismology,
gyrochronology, and isochrone fitting found consistent results.
Ongoing work by J.~Curtis et al.~broadly finds that the two
planet-hosting stars are indeed members of this diffuse population.
Given the cluster-level and individual-star evidence, this makes
Kepler-968 and Kepler-52 the youngest multiplanet systems currently
known from the Kepler primary mission.

@@

In the entire sample, this yielded \allagesyesconsistent\ consistent
cases, \allagesmaybeconsistent\ ``maybe'' cases, and
\allagesnoconsistent\ inconsistent.  Restricting our attention to the
$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
the respective counts are \minageltonegyryesconsistent,
\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
Rephrased, in the overall sample, \fracconsistentallages\% of stars
have consistent rotation and lithium-based ages
(\fracpotentiallyconsistentallages\% potentially consistent).  In the
sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
potentially consistent).  We feel it necessary to distinguish
``maybe'' consistent cases because of lingering uncertainties in how
the dispersion of the lithium data evolve over the first gigayear.  In
Appendix~\ref{app:inconsistent}, we discuss the
\minageltonegyrnoconsistent\ discrepant ``sub-gigayear'' stars in detail.
{\bf FIXME BRIEF SUMMARY, e.g. N of them are because lithium is garbage,
	M are because rotation is weird}.

@@

\appendix
\section{Cases With Inconsistent Rotation and Lithium-Based Ages}
\label{app:inconsistent}

For Kepler-957, $T_{\rm eff}$=4947\,K, and $P_{\rm rot}$=4.52\,days,
This implies a rotation-based age of $96_{-72}^{+192}$\,Myr.  The
rotational broadening of the spectrum ($v\sin
i$$\approx$9.8$\pm$1.0\kms) is consistent with the nominal stellar
rotation period.  However, if this star were $\lesssim$500\,Myr old,
lithium should have been detectable at $>$50\,m\AA.  Lithium is
marginally if at all present in the spectrum ({\bf  EW$_{\rm
Li}$$<$XX\,m\AA at 2$\sigma$ }).  While the planet (nominally
6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
\citep{2016ApJ...822...86M}, inspecting the data validation reports this is
somewhat surprising.  The light curve shows greater scatter in-transit
than out of transit, appears near-grazing, and has a suggestion of a
strong odd-even depth difference.  This is a surprising system that
merits further attention.

% from inconsitent_tli_tgyro cases:
Kepler-1644 b
Kepler-1699 b
 Kepler-786 b

these have finite Li ages, but no rotation age (perhaps b/c of the missing prots...)

Kepler-1943 b

 Kepler-101 b

\paragraph{Kepler-108 b and Kepler-108 c}---This sytem of two mutually
inclined giant planets \citep{2017AJ....153...45M} has a host star
that is entering the red giant branch.  This has been previously noted
through asteroseismology \citep{2013ApJ...767..127H}, and an analysis
of the Keck/HIRES spectrum with \texttt{SpecMatch-Synth} concurs, with
$T_{\rm eff}$=5668$\pm$100\,K, $\log g$=3.8$\pm$0.1,
[Fe/H]=0.38$\pm$0.06, and $v\sin i$=3.2$\pm$1.0\kms.

  Kepler-63 b

Kepler-1312 b Kepler-1312 c

   Kepler-1 b

K02680.01
K02639.01

@@

Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.

Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
The rotation-based age we find here is higher/lower than previously noted.

@@

Kepler~108 was reported by \citet{2020ApJ...903...55P} to be a member
of NGC6811 based ``on APOGEE radial velocities and Gaia-DR2 proper
motions''.  The cluster has $\mu_\alpha\approx-3.4$mas/yr and
$\mu_\delta \approx -8.8$mas/yr.  Kepler~108 has
$\mu_\alpha=-1.5\pm0.1$mas/yr and $\mu_\delta \approx
-1.2\pm0.1$mas/yr \citep{Gaia_DR3_2022}.  This is quite discrepant.

@@

\section{Comparison Against Previous Literature Age Scales}
Searching the literature for gyrochronal analyses of the Kepler field,
the most relevant studies seemed to be those of
\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
\citet{David_2021}.

Also Mathur2023, which does ``magneto-gyrochronology'', including the
Sph indicator in a model of the ages.

\subsection{Asteroseismic Age Comparison}
T Ceillier, J Van Saders et al 2016 MNRAS...

>>>>

\paragraph{Cep-Her}
The $\approx$40\,Myr Cep-Her complex currently has its most
constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs; it contains Kepler-1627Ab,
Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
\citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
may vary in age by $\approx$50\% (\citealt{Bouma_2022b};
Kerr~et.al.~in preparation).  The planets themselves are all on
close-in orbits (5-25\,days), with sizes from 2-4\,$R_\oplus$.  The
rotation, lithium, and cluster ages all agree for
Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported by \citet{Bouma_2022b} for the RSG-5 subgroup.

\paragraph{MELANGE-3}
\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
the Kepler field and derived its age using rotation,
lithium, and the isochronal pre-main-sequence.  Kepler-970 and
Kepler-1928 harbor two transiting mini-Neptunes inside this
association \citep{2022AJ....164...88B}.  Our rotation-based
age-dating independently recovers these systems, with $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based age
being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.

\paragraph{Theia~520}
We derive rotation-based ages of 330-350\,Myr for Kepler-968 and Kepler-52.
These stars were reported by
\citet{2019AJ....158..122K} to be members of an association called Theia~520
based on their spatial and kinematic correlations with nearby stars.
The core of this structure is also known as UBC-1
\citep{2018A&A...618A..59C}; since both Kepler planet hosts
are in diffuse regions of the cluster, we use the name from
\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting, Theia~520
is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
An independent analysis by
\citet{2024A&A...681A..13F} using g-mode asteroseismology,
gyrochronology, and isochrone fitting found consistent results.
Ongoing work by J.~Curtis et al.~broadly finds that the two
planet-hosting stars are indeed members of the diffuse population.
Given the cluster-level and individual-star evidence, this makes
Kepler-968 and Kepler-52 the youngest multiplanet systems currently
known from the primary Kepler mission.

@@

In the entire sample (i.e.~without imposing an quality cuts)
this yielded \allagesyesconsistent\ consistent
cases, \allagesmaybeconsistent\ ``maybe consistent'' cases, and
\allagesnoconsistent\ inconsistent.  Restricting our attention to the
$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
the respective counts are \minageltonegyryesconsistent,
\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
Rephrased, in the overall sample, \fracconsistentallages\% of stars
have consistent rotation and lithium-based ages
(\fracpotentiallyconsistentallages\% potentially consistent).  In the
sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
potentially consistent).  We feel it necessary to distinguish
``maybe'' consistent cases because of lingering uncertainties in how
the dispersion of the lithium data evolve over the first gigayear.

In
Appendix~\ref{app:inconsistent}, we discuss the
systems for which at least one age indicator
reports a sub-gigayear age in detail.
All but one are simply cases for which the star
is either evolved or else an unresolved binary.
The one exception is Kepler-786 (KOI-1199), an early K dwarf with 
$t_{\rm Li}$=\kepseveneightsix\,Myr, which would predict an obvious rotation signal.
No photometric signal is detected; other age indicators in the
spectrum similarly suggest that the star may not in fact be young.

{\bf FIXME BRIEF SUMMARY, e.g. N of them are because lithium is garbage,
	M are because rotation is weird}.

@@

\appendix
\section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
\label{app:inconsistent}

In this section we discuss fifteen systems with confirmed planets that have nominally
discrepant rotation and lithium-based ages.
We find that only one of them, Kepler-786, is interesting.

 \paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O}
has a G0 V primary and a $\approx$2.5\,day near-grazing transiting hot Jupiter,
and has been studied in detail by multiple investigators
\citep[e.g.][]{2007ApJ...664.1190S,2008ApJ...682.1283W,2011ApJ...733...36K,2011MNRAS.417.2166S}.
Our lithium-based age of this system
(\trestwotli\,Myr), derive from EW$_{\rm Li}$=81$\pm$3\,m\AA,
is qualitatively in line with the previously noted lithium abundance
\citep{2007ApJ...664.1190S}. 
However, we detect no photometric rotation signal, and the spectroscopic $v\sin i$ is low.
The lithium age is strongly asymmetric because
there is significant amounts of scatter in EW$_{\rm Li}$ vs.~$T_{\rm eff}$ over ages $t$$\gtrsim$1\,Gyr for early G dwarfs.
We calculate a $2\sigma$ upper limit on $t_{\rm Li}$ of 2.8\,Gyr, and a $3\sigma$ upper limit of 6.1\,Gyr.
The non-detection of rotation, while mildly surprising, is not shocking if the system's true
age is in this long tail.

\paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm Li}$=100$\pm$5\,m\AA,
nominally implying $t_{\rm Li}$$\approx$200-800\,Myr, but no rotation detection.
Our automated flags noted that this star has a low surface gravity, high luminosity, and is far from the main sequence.
\citet{2014A&A...572A...2B} concur: this star is a sub-giant slightly more massive than the Sun;
no rotation detection is expected; the lithium likely simply survived over the star's
main sequence lifetime.

\paragraph{Kepler-108}---This system of two mutually
inclined giant planets \citep{2017AJ....153...45M} has a $T_{\rm eff}$$\approx$5600\,K
host star with a strong lithium detection, and no rotation detection.
The host star is entering the red giant branch; this has been previously noted
through asteroseismology \citep{2013ApJ...767..127H}, and our analysis
of the Keck/HIRES spectrum with \texttt{SpecMatch-Synth} \citep{2017AJ....154..107P} concurs, yielding
$T_{\rm eff}$=5668$\pm$100\,K, $\log g$=3.8$\pm$0.1,
[Fe/H]=0.38$\pm$0.06, and $v\sin i$=3.2$\pm$1.0\kms.
Given the star's high mass, the high lithium content and lack of rotation detection
is not surprising.

%Independently, Kepler~108 was also reported by \citet{2020ApJ...903...55P} to be a member
%of NGC6811.  The cluster has $\mu_\alpha\approx-3.4$mas/yr and
%$\mu_\delta \approx -8.8$mas/yr.  Kepler~108 has
%$\mu_\alpha=-1.5\pm0.1$\,mas/yr and $\mu_\delta \approx
%-1.2\pm0.1$\,mas/yr \citep{Gaia_DR3_2022}.  We therefore suggest that it is not a member
%of NGC6811.

%
%Our quality flags noted that this star has
%a low surface gravity, a crowded field, high RUWE, and an ambiguous Kepler-to-Gaia crossmatch.

\paragraph{Kepler-786}---This K3 V star has a surprisingly 
high lithium content.
The spectrum is single-lined, with
 \texttt{SpecMatch-Synth} derived parameters of
 $T_{\rm eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1,
 [Fe/H]=0.11$\pm$0.06, and $v\sin i$=2.4$\pm$0.6\kms.
Yet with 
EW$_{\rm Li}$=83$\pm$6\,m\AA,
the lithium age of \kepseveneightsix\,Myr would predict an obvious rotation signal.
None is present in the Kepler light curve, consistent with the low
$v\sin i$.
Out of all ``apparently discrepant'' lithium and rotation measurements
discussed in this appendix, this is the only one that seems to remain discrepant after scrutiny.
The  \ion{Ca}{2} doublet is in emission, with $R'_{\rm HK}=-4.7\pm0.5$, which
suggests an age at least as old as the Hyades
\citep{2008ApJ...687.1264M}.
The Balmer lines are in absorption, and display no obvious signatures of youth.

\paragraph{Kepler-1644}---The rotation-based age of this system (\kepsixteenfourfour\,Myr) is nominally
much lower than the lithium limit ($>$767\,Myr).
The Kepler light curve shows a $\approx$1\% amplitude 1.4\,day rotation signal with many flares.
However the automated quality flags note that the star has low (photometric) surface gravity, a high RUWE (${\rm RUWE}_{\rm DR3}$=7.0),
and that it is far from the main sequence.
The Keck/HIRES spectrum also shows visually narrow lines, with a
 \texttt{SpecMatch-Synth} $v \sin i \leq 2$\kms. 
However, we performed a cross-correlation between the HIRES spectrum and 
the nearest matches in the Keck/HIRES template library \citep{2015AJ....149...18K}, and found that on 31 July 2022 (UT)
the system displayed
a broad CCF with a blended second component at $\approx$$+26$\,\kms relative to
the primary, with a best-fit flux-ratio of $\approx$3.4\%, and a preferred $T_{\rm eff,B}$$\approx$4400\,K.
The spectrum and astrometric excess noise therefore point to this system being an unresolved
binary, which calls the
reliability of the rotation-based age into question.
The non-detection of the companion
from Robo-AO imaging at Palomar \citep{2017AJ....153...66Z} suggests that the companion(s)
are likely within $\rho$$\lesssim$0.3$''$.

\paragraph{Kepler-1699}---This system is in a similar qualitative regime as Kepler-1644, with an
apparently young $t_{\rm gyro}$=\kepsixteenninenine\,Myr derived from a $\approx$2\% amplitude
4.2\,day rotation signal, and no evidence for lithium.
The system has ${\rm RUWE}_{\rm DR3}$=19.1.
From the same style of CCF analysis from the Keck/HIRES spectrum acquired on
31 Aug 2022 (UT), we also find a double-peaked CCF, in this case with a secondary
component at $\approx$$-16$\,\kms relative to
the primary with $T_{\rm eff,B}$$\approx$4900\,K.
This putative companion is similarly not detected in high-resolution imaging
 \citep{2017AJ....153...66Z}.
 The rotation-based age is questionable, given that we do not know which source the
 rotation signal is from, or whether the stars have interacted.

\paragraph{Kepler-1943}---This system nominally has a \kepnineteenfourthree\,Myr
lithium age, and no reported rotation detection.
However, the star is flagged as being over-luminous, low-surface gravity,
and far from the main sequence.
In other words, it is a subgiant.

\paragraph{Kepler-639, Kepler-320, Kepler-1719, Kepler-1876, Kepler-1072, Kepler-1743, Kepler-1929, Kepler-1488}---These eight systems all have nominally two-sided lithium age posteriors between 1 and 3\,Gyr, and yet lack rotation period detections.
All are subgiants, flagged with $Q_{\rm gyro}$ under various combinations of bits 1, 2, and 9.

@@

%Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
%
%Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
%The rotation-based age we find here is higher/lower than previously noted.

@@

@@

%\section{Comparison Against Previous Literature Age Scales}
%Searching the literature for gyrochronal analyses of the Kepler field,
%the most relevant studies seemed to be those of
%\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
%\citet{David_2021}.
%
%Also Mathur2023, which does ``magneto-gyrochronology'', including the
%Sph indicator in a model of the ages.
%
%
%\subsection{Asteroseismic Age Comparison}
%T Ceillier, J Van Saders et al 2016 MNRAS...
